---
output: html_document
editor_options: 
  chunk_output_type: console
---
# The basics: t-tests, ANOVA, and linear regression

The most basic statistical analyses for a GRA to use are t-tests, analysis of variance, and linear regression. Here are some tips and references for getting started and creating well-formatted output: 

## t-tests & ANOVA

The best R-package I have used for doing t-tests and ANOVA is [gtsummary](https://www.danieldsjoberg.com/gtsummary/), a package that makes beautiful tables for summarizing data. 

Here is an example of a t-test with `gtsummary`: 

```{r}
library(gtsummary)
# lets use the ToothGrowth data
str(ToothGrowth)

# Note: I am assuming here that you have checked the appropriate diagnostics (
#   e.g., check residuals for normality, check for outliers, check variances in both groups, etc.)

tbl_summary(ToothGrowth, # data set 
            by = "supp", # factor/grouping variable
            include = "len", # numeric outcome 
            label = list("len" ~ "Length"), # make a readable label 
            # t-test is a test of means, so let's summarize data with mean (sd) form
            statistic = list(all_continuous() ~ "{mean} ({sd})"), 
            # choose reasonable number of digits to print 
            digits = list(everything() ~ 2)) |> 
  # indicate that you want a t.test 
  add_p(test = list(all_continuous() ~ "t.test")) |> 
  # format labels 
  bold_labels() |> 
  # remove default 'Characteristic' header 
  modify_header(label = "")
```

For ANOVA, we could do this: 

```{r}
tbl_summary(ToothGrowth, 
            by = dose, # factor/grouping variable is now 'dose' 
            include = "len", 
            label = list("len" ~ "Length"), 
            statistic = list(all_continuous() ~ "{mean} ({sd})"), 
            digits = list(everything() ~ 2)) |> 
  # indicate that you want a one-way ANOVA test here, not assuming equal variances 
  add_p(test = list(all_continuous() ~ "oneway.test")) |> 
  bold_labels() |> 
  modify_header(label = "")
```


## Linear regression

`gtsummary` will also create summary tables for linear regression, like so: 

```{r}
# use the 'iris' data this time 
str(iris)
my_model <- lm(Petal.Length ~ Species + Sepal.Length, data = iris)

# again, I am assuming you have checked the appropriate linear model diagnostics
tbl_regression(my_model) |> 
  modify_header(label = "**Feature**", estimate = "**Estimate**") |> 
  bold_labels()
```


For logistic regression, you will need to do some futher customization of the output to get the odds ratio estimates to appear in the table - take a look at [this help documentation](https://www.danieldsjoberg.com/gtsummary/articles/tbl_regression.html) for examples. 

For more on diagnostics/checking assumptions: this article on [STHDA](http://www.sthda.com/english/articles/39-regression-model-diagnostics/161-linear-regression-assumptions-and-diagnostics-in-r-essentials/) gives some examples of checking assumptions for linear regression in `R`. 
