# Geographic (GIS) data analysis 

One really useful way to illustrate data is through making maps. Here's an example based on a real project I did in 2022 (data are available in the `data/` subfolder of this repository).

Suppose that a principal investigator (PI) wants to study access to care through teledentistry. The specific research question is something like, "in a population of patients treated at my dental clinic from January 2021 - May 2023, what were the patterns in the relationships between 1) the distance traveled to the clinic, 2) the mode of the intake exam (in-person or virtual), and 3) treatment completion within 6 months (yes or no). Suppose further that I (the analyst) have access to the county in which each person lives, in addition to the clinical information relevant to our research in the electronic dental records. 

Let's start by loading all the libraries we'll need: 

```{r libs, message=FALSE}
library(dplyr) # package for 'wrangling' data
library(sf) # package for analyzing GIS data 
library(ggplot2) # package for drawing plots/graphs
library(viridis) # package for choosing good color palettes 
library(knitr) # package for making tables
```

After loading libraries, the first thing I would do is look online to find [publicly available data](https://geodata.iowa.gov/datasets/iowa::iowa-county-boundaries/explore?location=42.011193%2C-93.543500%2C8.00) with [Federal Information Processing Standard (FIPS) codes](https://en.wikipedia.org/wiki/FIPS_county_code) for Iowa at the county level. I will use the R package `sf` to handle GIS data, which often comes in `.shp` or `GeoJSON` formats. Data in these sorts of files have what we need to draw a map of Iowa with the counties demarcated on the map. 

```{r}
# read in GeoJSO file 
ia <- st_read("data/Iowa_County_Boundaries.geojson")
# must create a 'geometry' object 
ia_geom <- st_geometry(ia)

# Note: in 'ia', O'Brien county is labeled as 'Obrien'-- as is often the case,
# special characters like apostraphes need special handling 
ia |> dplyr::filter(substr(CountyName,1,1) == "O")
# This is important - we will need to adjust our data later on to make things match.  

# But first, let's create a test plot to make sure the map looks right. 
par(mar = c(0,0,1,0))
plot(ia[1], reset = FALSE) # reset = FALSE: we want to add to a plot with a legend
```

This sample map is the correct shape -- here, the 99 counties are colored according to their index (1-99). We want to make a map where the colors correspond to the number of patients representing each county. 


Supposing one has access to electronic dental records, the data for such a study may look like this: 

```{r}
# read in CSV file that gives me county names
key <- read.csv("data/ia_counties.csv") |> 
  # narrow down to only IA (for sake of example)
  dplyr::filter(State == "IA") |> 
  # address the issue with the special character in O'Brien
  dplyr::mutate(NAME = dplyr::if_else(NAME == "O'Brien", "Obrien", NAME))
   
# now simulate mock data for our example 
# note: TD = teledentistry
set.seed(52242) 
td <- data.frame(
  # creating data from 1000 pretend IDs
  id = 1:1000,
  # randomly choose a year for each ID
  year = sample(2021:2023, 1000, replace = TRUE),
  # randomly choose an outcome for each ID 
  # (complete or incomplete, represented as 1 or 0)
  complete = sample(0:1, 1000, replace = TRUE),
  # randomly choose a county for each ID 
  county = sample(key$NAME, 1000, replace = TRUE),
  # randomly choose a model for each ID 
  mode = sample(c("virtual", "in-person"), 1000, replace = TRUE)
)

# see what these data look like: 
head(td) |>
  # note the use of the 'kable()' function from the 'knitr' package below: 
  # this is another great tool for making tables, when you want to see a 
  # glimpse of the data
  knitr::kable(caption = "Simulated data from electronic dental records")
```

Now, I am ready to create a map that will communicate to my collaborators where our patients are driving from to receive their treatment. 

```{r}
# determine how many patients in each county
td_summarize <- td |>
    group_by(county) |>
    summarise(N = n()) |>
  ungroup()

# add fips codes (from GeoJSON file)
td_summarize <- right_join(td_summarize, ia,
            by = c("county" = "CountyName")) 

# create sf object (for drawing a map)
map <- td_summarize |> 
  st_as_sf()

# draw the map 
ggplot() + 
  geom_sf(data = map,
          aes(fill = N)) + 
  scale_fill_viridis() + 
  labs(title = "Map of IA Residents by County", fill = "Number of \npatients") +
  theme_bw() 

```
